---
title: "Metadata checks"
author: "Kimberly Olney, Ph.D."
date: "07/18/2024"
output:
  html_document:
    df_print: paged
  pdf_document: default
params:
  args: myarg
---

Bulk RNAseq gene expression data used in this study is from the published manuscript "Distinct transcriptional alterations distinguish Lewy body disease from Alzheimer’s disease" by Olney et al. 2024, Brain\
Counts data is available on Synapse: syn52394100\

# Setup
```{r setup}
knitr::opts_knit$set(root.dir = ".")
library(dplyr)
library(data.table)
```

# Metadata
```{r metadata}
# Read in metadata file from the gene expression study
metadata <- read.delim("../metadata/expression_metadata.txt")

# Create an IID column using the NPID sample IDs
metadata$IID <- ifelse(
  grepl("NA", metadata$NPID),
  gsub("NA", "MC", gsub("-", "", metadata$NPID)),
  paste0("MC", gsub("-", "", metadata$NPID))
)

# Create a sex column; sex (1 = male, 2 = female)
metadata <- metadata %>%
  mutate(Sex_RNA = ifelse(sex_inferred == "male", 1, 2))
```

# Fam file
A .fam file consists of 6 columns where each row describes one sample:\
1. Family ID (FID)\
2. Individual ID (IID)\
3. Maternal ID (MID)\
4. Paternal ID (PID)\
5. Sex (1=Male,2=Female,-9=Unknown)\
6. phenotype (1=Control, 2=Case, -9=Unknown)\
```{r fam}
# read in .fam data
fam <- fread("../snp_array/632_CWOW.fam")
#fam_imputed <- fread("../snp_array/CWOW_TOPMED_Rsq08_QC.fam")

# Rename columns
fam <- fam %>%
  dplyr::rename(
    FID = V1,
    IID = V2, 
    MID = V3,
    PID = V4, 
    Sex = V5,
    Pheno = V6
  )
```

# Check data 
What samples are unique to either data set?
Does the sex match?
```{r check_data}
# Which IIDs are unique to either set? 
unqiue_fam <- setdiff(fam$IID, metadata$IID)
unqiue_meta <- setdiff(metadata$IID, fam$IID)

# Merge .fam data with metadata
fam_meta <- merge(fam, metadata, by = "IID")

# Check if the sex in fam match the inferred sex from the the RNA
matches <- fam_meta$Sex.x == fam_meta$Sex_RNA
table(matches) # There are three mis-matches
# Insepect the mismatches
non_matching_rows <- fam_meta[!matches, ]
# These non matches are the three that we identified to be incorrectly sex annotated in the RNA expression. 
# The plink sex check says that are correctly annotated, so we will remove so as to not cause downstream issues.
filter_meta <- subset(fam_meta, !(IID %in% non_matching_rows$IID))
```

# Investigate related pairs
Identity by descent analysis (IBD) was performed on the original (un-imputed) files and we identified 3 pairs of individuals that are potentially related PI_HAT > 0.25.\
For any future analysis we recommend excluding an individual from each of the following pairs.

MC10078-MC05057
MC13264-MC17200
MC07261-MC09314

```{r related_pairs}
# There are related pairs. We will add a sample from each pair to the exclude list
#MC10078-MC05057, exclude MC10078 (LBD) since MC05057 is AD. Also, MC10078 has poor RIN
#MC13264-MC17200
#MC07261-MC09314, exclude MC07261 since MC09314 is a clean control 
related_pairs <- c("MC05057", "MC10078", "MC13264", "MC17200", "MC07261", "MC09314")
# setdiff(related_pairs, metadata$IID)
# MC13264 not in metadata 

# related pairs metadata
related_pairs_meta <- subset(filter_meta, IID %in% related_pairs)

# related_pairs exclude list
ex_related <- c("MC10078", "MC07261")
```

# Exclude IIDs
```{r exclude_IDs}
# Write a table of the IIDs that are unique to .fam 
# and the three sex mix-matches to exclude
# Add a sample from each related pair 
exclude_fam_IID <- union(unqiue_fam, non_matching_rows$IID)
exclude_IID <- union(exclude_fam_IID, ex_related)
# Output the list of exclude IIDs that we will exclude downstream
# The file needs both fam ID and individual ID
exclude_IID_df <- as.data.frame(exclude_IID)
exclude_IID_df$FID <- exclude_IID_df$exclude_IID # Fam ID same as IID
# since fam is same as IID we don't need to reorder the columns
write.table(exclude_IID_df, "../metadata/exclude_fam_IID.txt", row.names = FALSE, quote = FALSE, col.names = FALSE)

# Now in the imputed files the fam is 0
# write an exclude table with fam 0 
exclude_IID_df$FID <- NULL
exclude_IID_df$FID <- c(0)

exclude_IID_df_Fam_first <- exclude_IID_df[, c("FID", "exclude_IID")]

write.table(exclude_IID_df_Fam_first, "../metadata/exclude_fam_IID_for_imputed_files.txt", row.names = FALSE, quote = FALSE, col.names = FALSE)


# Output the metadata containing both the fam and RNA metadata
# Samples in the filtered metadata are samples that have both RNA data and SNP data
filtered_meta <- subset(filter_meta, !(IID %in% exclude_IID))
filtered_fam <- subset(fam, IID %in% filtered_meta$IID)

write.table(filtered_meta, "../metadata/filtered_metadata.txt", row.names = FALSE, quote = FALSE, sep = "\t")
```


