---
title: "PlinkQC with LBD CWOW TOPMed-imputed SNP data"
author: "Kimberly Olney, Ph.D."
date: "March 2026"
output:
  html_document:
    theme: cerulean
    toc: true
    toc_float: true
  pdf_document:
    toc: true
editor_options:
  chunk_output_type: inline
---

# PlinkQC

PlinkQC is an R package for genotype quality control in genetic association studies. See https://meyer-lab-cshl.github.io/plinkQC/index.html

The protocol is implemented in three main functions:  
1) The per-individual quality control (`perIndividualQC`)  
2) The per-marker quality control (`perMarkerQC`)  
3) The generation of the new, quality control dataset (`cleanData`)  

Individuals and markers that fail quality control can subsequently be removed with plinkQC to generate a new, clean dataset.

```{r setup, echo=FALSE, warning=FALSE, message=FALSE}
.libPaths(c("/home/kolney/R/x86_64-pc-linux-gnu-library/"))
knitr::opts_knit$set(root.dir = ".")
library(plinkQC)
library(ggplot2)
Sys.setenv(RSTUDIO_PANDOC="/usr/local/biotools/pandoc/3.1.2/bin")
setwd("/tgen_labs/jfryer/kolney/LBD_CWOW/QTL/LBD_CWOW_QTL/")
```

```{r}
# Path to plinkQC package
package.dir <- find.package("plinkQC")

# Path to PLINK 1.9
path2plink <- "/tgen_labs/jfryer/kolney/tools/miniconda3/envs/QTL/bin/plink"

# Input directory containing PLINK bed/bim/fam files
indir <- "/tgen_labs/jfryer/kolney/LBD_CWOW/QTL/LBD_CWOW_QTL/snp_array/TOPMED_imput/imputed_vcf"

# Output directory for plinkQC plots and fail lists
qcdir <- "/tgen_labs/jfryer/kolney/LBD_CWOW/QTL/LBD_CWOW_QTL/results/plinkQC_plots_post_imputation"
dir.create(qcdir, recursive = TRUE, showWarnings = FALSE)

# Main CWOW dataset for sample-level QC
# This file should already have:
# - corrected FID/IID
# - sex and phenotype added
# - PAR split on chrX
# - heterozygosity-fail sample removed
name <- "CWOW_qc_ids_sex_pheno_splitPAR_noHetFail"

# Merged CWOW + HapMap dataset for ancestry evaluation
prefixMergedDataset <- "merge_HAP_CWOW_pca_final"
```

# Input files used in this Rmd
The following files are expected in indir:
CWOW_qc_ids_sex_pheno_splitPAR_noHetFail.bed
CWOW_qc_ids_sex_pheno_splitPAR_noHetFail.bim
CWOW_qc_ids_sex_pheno_splitPAR_noHetFail.fam

for sex check, heterozygosity/missingness, relatedness, and per-marker QC.
The ancestry check additionally uses the merged reference dataset:
merge_HAP_CWOW_pca_final.bed
merge_HAP_CWOW_pca_final.bim
merge_HAP_CWOW_pca_final.fam


### Sex check
check_sex identifies individuals with discordant reported and genetically inferred sex. This dataset was prepared with --split-par b38 prior to conversion to PLINK1 format so that chromosome X is handled appropriately.
```{r}
fail_sex <- check_sex(
  indir = indir,
  qcdir = qcdir,
  name = name,
  interactive = TRUE,
  verbose = FALSE,
  path2plink = path2plink,
  legend_text_size = 10,
  legend_title_size = 10,
  axis_text_size = 10,
  axis_title_size = 10,
  title_size = 12
)

ggsave(
  filename = file.path(qcdir, "sex_check_post_imputation.pdf"),
  width = 6,
  height = 5
)
```

### Heterozygosity and sample missingness check
The identification of individuals with outlying missing genotype and/or heterozygosity rates helps detect samples with poor DNA quality and/or concentration that should be excluded from the study. Typically, individuals with more than 3–7% of their genotype calls missing are removed. Outlying heterozygosity rates are judged relative to the overall heterozygosity rates in the study, and individuals whose rates are more than a few standard deviations from the mean heterozygosity rate are removed.
```{r}
fail_het_imiss <- check_het_and_miss(
  indir = indir,
  qcdir = qcdir,
  name = name,
  interactive = TRUE,
  verbose = FALSE,
  path2plink = path2plink,
  legend_text_size = 10,
  legend_title_size = 10,
  axis_text_size = 10,
  axis_title_size = 10,
  title_size = 12
)

ggsave(
  filename = file.path(qcdir, "heterozygosity_check_post_imputation.pdf"),
  width = 7,
  height = 6
)
```

### Relatedness check
Related individuals can be identified by their proportion of shared alleles at the genotyped markers (identity by descent, IBD). Standardly, individuals with second-degree relatedness or higher are excluded.
```{r}
exclude_relatedness <- check_relatedness(
  indir = indir,
  qcdir = qcdir,
  name = name,
  interactive = TRUE,
  verbose = FALSE,
  path2plink = path2plink,
  legend_text_size = 10,
  legend_title_size = 10,
  axis_text_size = 10,
  axis_title_size = 10,
  title_size = 12
)

ggsave(
  filename = file.path(qcdir, "IBD_check_post_imputation.pdf"),
  width = 10,
  height = 6
)
```

### SNP missingness check
Markers with excessive missingness rate are removed as they are considered unreliable. Typically, thresholds for marker exclusion based on missingness range from 1%–5%.
```{r}
fail_snpmissing <- check_snp_missingness(
  indir = indir,
  qcdir = qcdir,
  name = name,
  interactive = TRUE,
  verbose = FALSE,
  path2plink = path2plink,
  showPlinkOutput = FALSE,
  legend_text_size = 10,
  legend_title_size = 10,
  axis_text_size = 10,
  axis_title_size = 10,
  title_size = 12
)

ggsave(
  filename = file.path(qcdir, "snp_missingness_check_post_imputation.pdf"),
  width = 10,
  height = 6
)
```

### HWE check
Markers with strong deviation from Hardy–Weinberg equilibrium (HWE) might indicate genotyping or genotype-calling errors. The HWE-filtered dataset used here has already undergone control-based HWE filtering upstream, but this section allows visual inspection of the distribution within plinkQC.
```{r}
fail_hwe <- check_hwe(
  indir = indir,
  qcdir = qcdir,
  name = name,
  interactive = TRUE,
  verbose = FALSE,
  path2plink = path2plink,
  showPlinkOutput = FALSE,
  legend_text_size = 10,
  legend_title_size = 10,
  axis_text_size = 10,
  axis_title_size = 10,
  title_size = 12
)

ggsave(
  filename = file.path(qcdir, "hwe_check_post_imputation.pdf"),
  width = 10,
  height = 5
)
```

### MAF check
Markers with low minor allele frequency (MAF) are often removed because genotype calling becomes more difficult as the heterozygote and rare-homozygote clusters become sparse. The current dataset has already undergone upstream MAC filtering, but this section provides a visual summary of the retained allele-frequency distribution.
```{r}
fail_maf <- check_maf(
  indir = indir,
  qcdir = qcdir,
  name = name,
  interactive = TRUE,
  verbose = FALSE,
  path2plink = path2plink,
  showPlinkOutput = FALSE,
  legend_text_size = 10,
  legend_title_size = 10,
  axis_text_size = 10,
  axis_title_size = 10,
  title_size = 12
)

ggsave(
  filename = file.path(qcdir, "maf_check_post_imputation.pdf"),
  width = 4.5,
  height = 5.5
)
```
# Per-marker QC summary
This section summarizes the marker-level QC outputs produced by plinkQC.
```{r}
fail_markers <- perMarkerQC(
  indir = indir,
  qcdir = qcdir,
  name = name,
  path2plink = path2plink,
  verbose = TRUE,
  interactive = TRUE,
  showPlinkOutput = FALSE
)

ggsave(
  filename = file.path(qcdir, "summary_check_post_imputation.pdf"),
  width = 8.5,
  height = 11
)
```

### Ancestry check
The identification of individuals of divergent ancestry is achieved by combining the study genotypes with a reference dataset consisting of individuals from known ancestral populations, here HapMapIII. Principal component analysis (PCA) on the combined genotype panel is used to detect population structure.
evaluate_check_ancestry uses principal components 1 and 2 to determine the center of the European reference samples. Study samples with a Euclidean distance from this center that exceeds a specified radius are classified as non-European.
The default radius is 1.5. For this dataset, a radius of 5 was selected to retain more individuals.

PCA via plink and LD prune
plink \
  --bfile merge_HAP_CWOW_pca_final \
  --pca 20 \
  --out merge_HAP_CWOW_pca_final

plink \
  --bfile merge_HAP_CWOW_pca_final \
  --indep-pairwise 200 50 0.2 \
  --out merge_HAP_CWOW_pca_final_pruned

plink \
  --bfile merge_HAP_CWOW_pca_final \
  --extract merge_HAP_CWOW_pca_final_pruned.prune.in \
  --pca 20 \
  --out merge_HAP_CWOW_pca_final

head CWOW_qc_ids_sex_pheno_splitPAR_noHetFail.renamed.sscore
plink --bfile merge_HAP_CWOW_pca_final --indep-pairwise 200 50 0.2 --out merge_HAP_CWOW_pca_final_pruned
plink --bfile merge_HAP_CWOW_pca_final --extract merge_HAP_CWOW_pca_final_pruned.prune.in --pca 20 --out merge_HAP_CWOW_pca_final
```{r}
path2plink2 <- "/tgen_labs/jfryer/kolney/tools/miniconda3/envs/QTL/bin/plink2"
path2load_mat <- "/tgen_labs/jfryer/kolney/tools/plinkQC/inst/extdata/load_mat/merged_chrs.postQC.train.pca"

file.exists(paste0(path2load_mat, ".acount"))
file.exists(paste0(path2load_mat, ".eigenvec.allele"))

exclude_ancestry <- ancestry_prediction(
  indir = indir,
  qcdir = qcdir,
  name = name,
  path2plink2 = path2plink2,
  path2load_mat = path2load_mat,
  interactive = TRUE,
  verbose = FALSE,
  legend_text_size = 10,
  legend_title_size = 10,
  axis_text_size = 10,
  axis_title_size = 10,
  title_size = 12,
  do.run_ancestry_prediction = TRUE,
  do.evaluate_ancestry_prediction = TRUE
)

ggsave(
  filename = file.path(qcdir, "ancestry_check_post_imputation.pdf"),
  width = 9,
  height = 6.5
)

# PCA
pca <- fread(file.path(qcdir, "CWOW_qc_ids_sex_pheno_splitPAR_noHetFail.renamed.sscore"))
head(pca)
ggplot(pca, aes(x = PC1_AVG, y = PC2_AVG)) +
  geom_point(size = 2, alpha = 0.8) +
  theme_bw() +
  labs(
    title = "Ancestry PCA projection",
    x = "PC1",
    y = "PC2",
    color = "Predicted ancestry"
  ) +
  theme(
    text = element_text(size = 12),
    plot.title = element_text(hjust = 0.5)
  )

library(data.table)
library(ggplot2)

pca <- fread("/tgen_labs/jfryer/kolney/LBD_CWOW/QTL/LBD_CWOW_QTL/snp_array/TOPMED_imput/imputed_vcf/merge_HAP_CWOW_pca_final.eigenvec")
colnames(pca)[1:4] <- c("FID", "IID", "PC1", "PC2")

pca$dataset <- ifelse(grepl("^MC", pca$IID), "CWOW", "HapMap")

ggplot(pca, aes(PC1, PC2, color = dataset)) +
  geom_point(alpha = 0.7, size = 2) +
  theme_bw() +
  labs(
    title = "CWOW + HapMap PCA",
    x = "PC1",
    y = "PC2",
    color = "Dataset"
  )

exclude_ancestry <- ancestry_prediction(
  indir = indir,
  qcdir = qcdir,
  name = name,
  path2plink2 = path2plink2,
  path2load_mat = path2load_mat,
  excludeAncestry = c("Africa", "America", "Central_South_Asia", "East_Asia", "Middle_East"),
  interactive = TRUE,
  verbose = FALSE,
  do.run_ancestry_prediction = TRUE,
  do.evaluate_ancestry_prediction = TRUE
)

write.table(
  exclude_ancestry$exclude_ancestry,
  file = file.path(qcdir, paste0(name, ".exclude-ancestry.IDs")),
  quote = FALSE,
  sep = "\t",
  row.names = FALSE,
  col.names = FALSE
)
```

# Output cleaned data
cleanData generates a cleaned dataset after removing all individuals and markers failing the plinkQC thresholds.
```{r}
Ids <- cleanData(
  indir = indir,
  qcdir = qcdir,
  name = name,
  path2plink = path2plink,
  verbose = TRUE,
  showPlinkOutput = FALSE
)
```

# No additional filtering was applied after plinkQC evaluation.
# The final dataset retained for downstream GWAS/eQTL analyses was:
```{r}
final_dataset <- file.path(indir, name)
final_dataset
```