---
title: "Process genotype and expression files for eQTL analysis"
author: "Kimberly Olney, Ph.D."
date: "March 2026"
output:
  html_document:
    df_print: paged
  pdf_document: default
params:
  args: myarg
---

This notebook prepares matched genotype, metadata, and expression matrices for eQTL analysis using the final cleaned 579-sample imputed genotype dataset.

The final genotype input is:CWOW_TOPMED_final_postQC_n579.clean_genotype.raw

The additive genotype file is read directly into R, subset to the final 579 samples, reordered to match metadata, and transposed so that SNPs are rows and samples are columns.

# Setup
```{r setup}
.libPaths(c("/home/kolney/R/x86_64-pc-linux-gnu-library/"))
knitr::opts_knit$set(root.dir = ".")
setwd("/tgen_labs/jfryer/kolney/LBD_CWOW/QTL/LBD_CWOW_QTL/scripts/")
source(here::here("/tgen_labs/jfryer/kolney/LBD_CWOW/QTL/LBD_CWOW_QTL/scripts/", "file_paths_and_colours.R"))

suppressPackageStartupMessages({
  library(data.table)
  library(duckdb)
  library(GenomicFeatures)
  library(edgeR)
  library(sva)
  library(DGEobj.utils)
  library(ggplot2)
  library(txdbmaker)
})
```

# Read in filtered metadata for the final 579-sample dataset
```{r}
# metadata
metadata <- fread("../metadata/n579_metadata.txt")
# genotype
genotype_matrix <- fread("../snp_array/final_gwas_dataset/CWOW_TOPMED_final_postQC_n579.genotype_formatted.txt")
genotype_matrix <- as.data.frame(genotype_matrix)

# first column should be SNP IDs after transpose
rownames(genotype_matrix) <- genotype_matrix[[1]]
genotype_matrix[[1]] <- NULL

# make sure sample columns match metadata
genotype_matrix <- genotype_matrix[, metadata$IID, drop = FALSE]
stopifnot(all(colnames(genotype_matrix) == metadata$IID))

# Save the formatted data to a file
write.table(genotype_matrix, file = "../snp_array/CWOW_n579_filtered_genotype_data.txt", sep = " ", quote = FALSE, col.names = NA)
```


# Gene length
Obtain summed exonic gene lengths for conversion of counts to TPM values.
```{r}
# Read in GTF and construct TxDb
txdb <- txdbmaker::makeTxDbFromGFF(
  "/tgen_labs/jfryer/projects/references/human/GRCh38/gencode.v38.annotation.sorted.gtf",
  format = "gtf"
)

# Collect exons per gene
exons.list.per.gene <- exonsBy(txdb, by = "gene")

# Reduce overlapping exons and sum widths
exonic.gene.sizes <- sum(width(reduce(exons.list.per.gene)))

# Save as data.frame
exonic_length <- as.data.frame(exonic.gene.sizes)
exonic_length$gene_id <- rownames(exonic_length)

# Read gene information used in the bulk RNA-seq data
genes <- fread("../RNA_counts/genes_filtered_protein_coding.txt")

# Add exon length information
genes_with_exonic_length <- merge(genes, exonic_length, by = "gene_id")

# Inspect
head(genes_with_exonic_length)
```

# Counts data setup
The bulk RNA-seq count matrix is subset to the 579 samples retained in the final genotype dataset, renamed from NPID to IID, and reordered to match metadata and genotype sample order.
```{r}
# Read counts table
counts <- fread("../RNA_counts/counts_filtered.txt")
counts <- as.data.table(counts)

# First column contains gene IDs
gene_ids <- counts[[1]]

# Keep only columns present in metadata NPID
sample_cols <- intersect(colnames(counts), metadata$NPID)
filtered_counts <- counts[, c(colnames(counts)[1], sample_cols), with = FALSE]

# Convert to data.frame and assign rownames
filtered_counts_df <- as.data.frame(filtered_counts)
rownames(filtered_counts_df) <- gene_ids
filtered_counts_df[[1]] <- NULL

# Create NPID -> IID mapping
matching_vector <- setNames(metadata$IID, metadata$NPID)

# Rename columns from NPID to IID
colnames(filtered_counts_df) <- matching_vector[colnames(filtered_counts_df)]

# Reorder columns to match metadata IID order
filtered_counts_df <- filtered_counts_df[, metadata$IID, drop = FALSE]

# Convert to numeric matrix
counts_matrix <- as.matrix(filtered_counts_df)
mode(counts_matrix) <- "numeric"

# Sanity checks
stopifnot(all(colnames(counts_matrix) == metadata$IID))
stopifnot(all(colnames(counts_matrix) == colnames(genotype_matrix)))

# Inspect
dim(counts_matrix)[1:2]
head(rownames(counts_matrix))
head(colnames(counts_matrix))
```

# ComBat-Seq batch correction
Apply batch correction to the count matrix while preserving disease group.
```{r}
adjusted <- ComBat_seq(
  counts = counts_matrix,
  batch = metadata$batch,
  group = metadata$TYPE
)
#Compare raw and adjusted counts

#This section compares log2 CPM values before and after ComBat-Seq correction on a per-sample basis.

lcpm <- edgeR::cpm(counts_matrix, log = TRUE)
ladjusted <- edgeR::cpm(adjusted, log = TRUE)

num_samples <- ncol(counts_matrix)
```

```{r plot, eval=FALSE}
for (i in seq_len(num_samples)) {
  lcpm_sample <- lcpm[, i]
  ladjusted_sample <- ladjusted[, i]
  corr_value <- round(cor(lcpm_sample, ladjusted_sample), 2)
  plot_title <- paste("Sample", colnames(lcpm)[i], "\nCorrelation:", corr_value)

  p <- ggplot(
    data.frame(lcpm_sample = lcpm_sample, ladjusted_sample = ladjusted_sample),
    aes(x = lcpm_sample, y = ladjusted_sample)
  ) +
    geom_point(alpha = 0.6) +
    geom_smooth(method = "lm", se = FALSE, color = "blue") +
    labs(
      title = plot_title,
      x = "Log2 CPM (Original)",
      y = "Log2 CPM (Adjusted)"
    ) +
    theme_minimal()

  plot(p)

  ggsave(
    paste0("../results/combat/", colnames(lcpm)[i], ".pdf"),
    width = 4,
    height = 4
  )
}
dev.off()
```


# TPM counts
Convert raw and ComBat-adjusted counts to log2 TPM.
```{r}
# TPM from ComBat-adjusted counts
adjusted_nozero <- adjusted
adjusted_nozero[adjusted_nozero == 0] <- 1e-6

TPM_counts_adjusted <- convertCounts(
  adjusted_nozero,
  unit = "TPM",
  genes_with_exonic_length$exonic.gene.sizes,
  log = TRUE
)

write.table(
  TPM_counts_adjusted,
  "../RNA_counts/n579_TPM_combat_adjusted_counts_log2.txt",
  sep = "\t",
  quote = FALSE
)

# TPM from raw counts
counts_nozero <- counts_matrix
counts_nozero[counts_nozero == 0] <- 1e-6

TPM_counts_raw <- convertCounts(
  counts_nozero,
  unit = "TPM",
  genes_with_exonic_length$exonic.gene.sizes,
  log = TRUE
)

write.table(
  TPM_counts_raw,
  "../RNA_counts/n579_TPM_counts_log2.txt",
  sep = "\t",
  quote = FALSE
)
```

Save matched outputs for eQTL analysis
Save matched metadata, genotype, and expression matrices for downstream eQTL workflows.

# Save matched metadata
```{r}
fwrite(
  metadata,
  "../metadata/n579_metadata_matched_for_eQTL.txt",
  sep = "\t"
)

# Save genotype matrix
fwrite(
  genotype_matrix,
  file = "../snp_array/final_gwas_dataset/CWOW_TOPMED_final_postQC_genotype_matrix.txt",
  sep = "\t",
  quote = FALSE,
  col.names = NA
)

# Save raw matched counts
write.table(
  counts_matrix,
  file = "../RNA_counts/n579_TPM_counts_matrix_matched.txt",
  sep = "\t",
  quote = FALSE,
  col.names = NA
)

# Save ComBat-adjusted counts
write.table(
  adjusted,
  file = "../RNA_counts/n579_TPM_combat_adjusted_counts.txt",
  sep = "\t",
  quote = FALSE,
  col.names = NA
)
```

# Final checks
Confirm that genotype, counts, and metadata all use the same 579 samples in the same order.
```{r}
cat("Metadata samples: ", nrow(metadata), "\n")
cat("Genotype samples: ", ncol(genotype_matrix), "\n")
cat("Counts samples: ", ncol(counts_matrix), "\n")

stopifnot(nrow(metadata) == ncol(genotype_matrix))
stopifnot(nrow(metadata) == ncol(counts_matrix))
stopifnot(all(metadata$IID == colnames(genotype_matrix)))
stopifnot(all(metadata$IID == colnames(counts_matrix)))

cat("All matched sample checks passed.\n")
```